Pervasive duplication of tumor suppressor genes preceded parallel evolution of large bodied Atlantogenatans

### 2020-05-20 paper_PLOS/paper_PLOS_draft.tex
first draft

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Pervasive duplication of tumor suppressor genes preceded parallel
evolution of large bodied Paenungulates} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Juan Manuel Vazquez\textsuperscript{\getAff{University of Chicago}}\textsuperscript{*},
Vincent J Lynch\textsuperscript{\getAff{University at Buffalo, SUNY}}\\
\bigskip
\textbf{\getAff{University of Chicago}}Department of Human Genetics, 920 East 58th St, Chicago, IL, 60637\\
\textbf{\getAff{University at Buffalo, SUNY}}Department of Biological Sciences, 551 Cooke Hall, Buffalo NY, 14260\\
\bigskip
* Corresponding author: juanvazquez@uchicago.edu\\
\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
Lorem ipsum dolor sit amet, consectetur adipiscing elit. Curabitur eget
porta erat. Morbi consectetur est vel gravida pretium. Suspendisse ut
dui eu ante cursus gravida non sed sem. Nullam sapien tellus, commodo id
velit id, eleifend volutpat quam. Phasellus mauris velit, dapibus
finibus elementum vel, pulvinar non tellus. Nunc pellentesque pretium
diam, quis maximus dolor faucibus id. Nunc convallis sodales ante, ut
ullamcorper est egestas vitae. Nam sit amet enim ultrices, ultrices elit
pulvinar, volutpat risus.

@@

\emph{Reciprocal Best-Hit BLAT:} We developed a reciprocal best hit BLAT
(RBHB) pipeline to quickly identify homologs and estimate gene copy
numbers (\textbf{Figure 1A}). The Reciprocal Best Hit (RBH) search
strategy is conceptually straightforward: 1) Given a gene of interest
\(G_A\) in a query genome \(A\), one searches a target genome \(B\) for
all possible matches to \(G_A\); 2) For each of these hits, one then
performs the reciprocal search in the original query genome to identify
the highest-scoring hit; 3) A hit in genome \(B\) is defined as a
homolog of gene \(G_A\) if and only if the original gene \(G_A\) is the
top reciprocal search hit in genome \(A\). We selected BLAT {[}26{]} as
our algorithm of choice, as this algorithm is sensitive to highly
simliar (\textgreater{}90\% identity) sequences, thus identifying the
highest-confidence homologs while minimizing many-to-one mapping
problems when searching for multiple genes. RBH performs similar to
other more complex methods of orthology prediction, and is particularly
good at identifying incomplete genes that may be fragmented in low
quality/poor assembled regions of the genome {[}{\textbf{???}},27{]}.

\emph{Effective Copy Number By Coverage:} In lower-quality genomes, many
genes are fragmented across multiple scaffolds, which results in BLAT
calling multiple hits when in reality there is only one gene. To
compensate for this, we came up with a novel statistic, Estimated Copy
Number by Coverage (ECNC), which averages the number of times we see
each nucleotides of a query sequence in a target genome over the total
number of nucleotides of the query sequence found overall in each target
genome (Supplementary Figure 1). This allows us to correct for genes
that have been fragmented across incomplete genomes, while also taking
into account missing sequences from the human query in the target
genome. Mathematically, this can be written as:\\
\[ ECNC = \frac{\sum_{n=1}^{l} C_n}{\sum_{n=1}^{l} bool(C_n)}\] where
\(n\) is a given nucleotide in the query, \(l\) is the total length of
the query, \(C_n\) is the number of instances that \(n\) is present
within a reciprocal best hit, and \(bool(C_n)\) is 1 if \(C_n > 0\) or 0
if \(C_n = 0\).

\emph{RecSearch Pipeline:} We created a custom Python pipeline for
automating RBHB searches between a single reference genome and multiple
target genomes using a list of query sequences from the reference
genome. For the query sequences in our search, we used the hg38 Proteome
provided by UniProt {[}28{]}, which is a comprehensive set of protein
sequences curated from a combination of predicted and validated protein
sequences generated by the UniProt Consortium. In order to refine our
search, we omitted protein sequences originating from long, noncoding
RNA loci (e.g.~LINC genes); poorly-studied genes from predicted open
reading frames (C-ORFs); and sequences with highly repetitive sequences
such as zinc fingers, protocadherins, and transposon-containing genes,
as these were prone to high levels of false positive hits. After
filtering out problematic protein queries, we then used our pipeline
(Figure 1A) to search for all copies of our 20456 query genes in
publicly available Afrotherian genomes, including African savannah
elephant (\emph{Loxodonta africana}: loxAfr3, loxAfr4, loxAfrC), African
forest elephant (\emph{Loxodonta cyclotis}: loxCycF), Asian Elephant
(\emph{Elephas maximus}: eleMaxD), Woolly Mammoth (\emph{Mammuthus
primigenius}: mamPriV), Colombian mammoth (\emph{Mammuthus columbi}:
mamColU), American mastodon (\emph{Mammut americanum}: mamAmeI), Rock
Hyrax (\emph{Procavia capensis}: proCap1, proCap2, proCap2\_HiC), West
Indian Manatee (\emph{Trichechus manatus latirostris}: triManLat1,
triManLat1\_HiC), Aardvark (\emph{Orycteropus afer}: oryAfe1,
oryAfe1\_HiC), Lesser Hedgehog Tenrec (\emph{Echinops telfairi}:
echTel2), Nine-banded armadillo (\emph{Dasypus novemcinctus}: dasNov3),
Hoffman's two-toed sloth (\emph{Choloepus hoffmannii}: choHof1, choHof2,
choHof2\_HiC), Cape golden mole (\emph{Chrysochloris asiatica}:
chrAsi1), and Cape elephant shrew (\emph{Elephantulus edwardii}:
eleEdw1). For many of these species, we covered multiple assemblies in
order to test the effects of assembly size and quality on our hits.

\emph{Duplication gene inclusion criteria:} In order to condense
transcript-level hits into single gene loci, and to resolve many-to-one
genome mappings, we removed exons where transcripts from different genes
overlapped, and merged overlapping transcripts of the same gene into a
single gene locus call. The resulting gene-level copy number table was
then combined with the maximum ECNC values observed for each gene in
order to call gene duplications. We called a gene duplicated if its copy
number was two or more, and if the maximum ECNC value of all the gene
transcripts searched was 1.5 or greater; previous studies have shown
that incomplete duplications can encode functional genes, therefore
partial gene duplications were included provided they passed additional
inclusion criteria. The ECNC cut off of 1.5 was selected empirically, as
this value minimized the number of false positives seen in a test set of
genes and genomes. The results of our initial search are summarized in
Figure 1B. Overall, we identified {[}MEDIAN{]} genes across all species,
or {[}\%HITS/QUERIES{]} of our starting query genes.

\emph{Duplicate gene exclusion criteria:} We excluded genes from
downstream analyses for which assignment of homology was uncertain,
including uncharacterized ORFs (17), LOC (17), HLA genes (17),
replication dependent histones (17), odorant receptors (17), ribosomal
proteins (17), zinc finger transcription factors (17), viral and
repetitive-element-associated proteins (17) and any protein described as
either ``Uncharacterized,'' ``Putative,'' or ``Fragment'' by UniProt in
UP000005640 (17).

\textbf{Orthogonal Genome Assessment using CEGMA} In order to determine
the effect of genome quality on our results, we used the gVolante
webserver and CEGMA to assess the quality and completeness of the
genome. CEGMA was run using the default settings of {[}{]}, and the
mammalian-specific core gene sets.

@@

In order to determine the cancer risk at each node, we used a simplified
multistage cancer risk model for body size and lifespan
{[}{\textbf{???}},{\textbf{??}},{\textbf{??}}{]}. We defined the

@@

\begin{figure}
\centering
\includegraphics{paper_PLOS_draft_files/figure-latex/Figure 1-1.pdf}
\caption{Figure 1: Body sizes rapidly and frequently expand in
Eutherians, especially in Atlantogenata. \textbf{A)} Tree of Eutherian
species, colored by ln(Body Size) and with branch lengths set to the
rate of change in body sizes, normalized by the square root of the root
branch. Atlantogenata is highlighted at the bottom. \textbf{B)} Zoom-in
of (A) on Atlantogenata. Silhuetes for the African Elephant, West Indian
Manatee, Cape Elephant Shrew, Lesser Hedgehog Tenrec, Cape Golden Mole,
Nine-Banded Armadillo, and Hoffman's Two-Toed Sloth are colored by their
extant body sizes, while clade labels are colored based on the common
ancestor's estimated body size}
\end{figure}

@@

Ancestor/Species & Estimated Body Size (log(g)) & 95\% CI (Low) & 95\%
CI (High) & Rate (sqrt)\tabularnewline

@@

To trace the evolutionary history of body mass and lifespan in
Afrotherians, we built a time-calibrated supertree of Eutherian mammals
combining 1,679 species from Bininda-Emonds et al {[}24{]} with a total
evidence Afrotherian phylogeny including 77 extant and fossil data from
39 extinct species {[}19{]}. Fossil data from extinct species were
included to ensure that ancestral state reconstructions of body mass in
Afrotherians were not biased by only including extant species, which can
lead to inaccurate reconstructions, for example, if lineages multiple
lineages evolved large body masses from a small bodied ancestor. We
jointly estimated rates of body mass evolution and reconstructed
ancestral states using a generalization of a Brownian model of character
evolution, which allows for occasional large jumps in traits (stable
model) and out performs standard Brownian motion and Ornstein?Uhlenbeck
models of character evolution {[}25{]}.

Similar to previous studies of Afrotherian body size {[}19,25{]}, we
found that the body mass of the Afrotherian ancestor was inferred to be
small (0.26kg, 95\% CI: 0.31-3.01kg) and that substantial accelerations
in the rate of body mass evolution occurred coincident with a 65×
increase in body mass in the stem-lineage of \emph{Pseudoungulata}
(17kg), a 1.5× increase in body mass in the stem-lineage of
\emph{Paenungulata} (25kg), a 12× increase in body mass in the
stem-lineage of \emph{Tehthytheria} (296kg), and a 14× increase in body
mass in the stem-lineage of \emph{Proboscidea} (4,100kg; Figure 1). The
ancestral \emph{Hyracoidea} was inferred to be relatively small
(2.86-15.71kg), and rate accelerations were coincident with independent
body mass increases in large hyraxes such as \emph{Titanohyrax andrewsi}
(67× increase in body mass). While the body mass of the ancestral
\emph{Sirenian} was inferred to be large (61-656kg), a rate acceleration
occurred coincident with a 10× body mass increase in Stellar's sea cow.
Rate accelerations also occurred coincident with 36× body mass reduction
in the stem-lineage of the dwarf elephants \emph{Elephas}
(\emph{Palaeoloxodon}) \emph{falconeri} and \emph{Palaeoloxodon
cypriotes}. These data suggest that gigantism in \emph{Afrotherians}
evolved step-wise, from small to medium bodies in the
\emph{Pseudoungulata} stem-lineage, medium to large bodies in the
\emph{Tehthytherian} stem-lineage and extinct hyraxes, and from large to
exceptionally large bodies independently in the \emph{Proboscidean}
stem-lineage and Stellar's sea cow (Figure 1).

\hypertarget{pervasive-duplication-of-tumor-suppressors-during-the-origins-of-large-bodied-afrotheirans}{%
\subsection{Pervasive duplication of tumor suppressors during the
origins of large bodied
Afrotheirans}\label{pervasive-duplication-of-tumor-suppressors-during-the-origins-of-large-bodied-afrotheirans}}

\includegraphics{paper_PLOS_draft_files/figure-latex/Figure 2C1-1.pdf}
\includegraphics{paper_PLOS_draft_files/figure-latex/Figure 2C2-1.pdf}

\begin{verbatim}
## Warning: TODO: Table 2 is done, but Stargazer has some weird bug stopping it
## from running...
\end{verbatim}

Enhanced cancer suppression may have evolved through many mechanisms;
among the most parsimonious is an increase in the copy number of genes
with tumor suppressor functions. Previous studies focusing on candidate
gene studies, for example, have identified increased copy number of the
tumor suppressors \emph{TP53} and \emph{LIF} in elephants
{[}{\textbf{???}},11,21--23{]}. Therefore, in order to test whether this
was a pervasive phenomena genome-wide in \emph{Afrotherians}, we used a
Reciprocal Best Hit BLAT (RBHB) approach to infer gene copy number in
\emph{Afrotherian} and \emph{Atlantogenatan} genomes (Fig. 2A). Because
RBHB-like approaches can over-estimate copy number when genes are
fragmented or incorrectly assembled across multiple scaffolds, we also
inferred copy number using a complementary method that quantifies the
ratio between observed and expected gene coverage per nucleotide (ECNC)
(Supplementary Figure 1). By only including nucleotides from the query
sequence that were observed in the target genome, we also correct for
partial hits where some or all of the homologs of a gene have diverged
from the human homolog.

Because our sequence database included various protein transcripts for
each gene, in order to obtain gene-level copy number information and
elimate any many-to-one mappings of hits, we labeled each exon of every
reciprocal best hit (RBH) with the gene corresponding to the query
transcript and merged all overlapping exons; next, we eliminated any
many-to-one exons that resulted from the previous step. Finally, we
reassembled the gene loci based on the original transcript starts and
ends, and the collapsed exon data, obtaining the full sequence of each
RBH locus. Genes were considered to be duplicated if its copy number via
RBHB was greater than or equal to 2, and the maximum ECNC among all
transcripts prior to filtering was greater than or equal to 1.50. This
cutoff of ECNC was selected to account for truncated gene duplications,
which have been shown to be functional in various examples
{[}{\textbf{???}} examples of this{]}.

To reconcile the Atlantogenatan phylogeny with duplication events, we
used maximum likelihood to reconstruct likely ancestral copy numbers for
each gene at each node in the phylogeny. To define the copy number of a
gene, we conservatively used the lesser value between the RBHB hit
count, and the ECNC value rounded to the nearest whole number. In order
to perform

Next, in order to select genes and duplicates which were likely
functional, we omitted any hits that were not supported by either the
gene prediction method GenScan, or by at least one transcript assembled
from publically-available RNA-seq data.

We describe the number of genes that increased in each lineage in
\emph{Atlantogenata} in Figure 2. Among the genes that increased in copy
number in the elephant lineage are TP53 and LIF, as previously
described. Furthermore, we identify

\hypertarget{duplications-that-occured-recently-in-probodiscea-are-enriched-for-tumor-suppressor-pathways}{%
\subsection{Duplications that occured recently in Probodiscea are
enriched for tumor suppressor
pathways}\label{duplications-that-occured-recently-in-probodiscea-are-enriched-for-tumor-suppressor-pathways}}

\includegraphics{paper_PLOS_draft_files/figure-latex/Fig 3-1.pdf}
\includegraphics{paper_PLOS_draft_files/figure-latex/Fig 3-2.pdf}

In order to infer the functional consequences of these gene
duplications, we tested if duplicate genes were enriched in specific
pathways relative to our initial query set of genes. We used

\hypertarget{concerted-duplication-of-tp53-and-tp53-related-genes-towards-probodiscea}{%
\subsection{\texorpdfstring{Concerted duplication of TP53 and
TP53-related genes towards
\emph{Probodiscea}}{Concerted duplication of TP53 and TP53-related genes towards Probodiscea}}\label{concerted-duplication-of-tp53-and-tp53-related-genes-towards-probodiscea}}

\begin{verbatim}
## Warning: TODO: Make Fig4
\end{verbatim}

\hypertarget{step-wise-reduction-of-intrinsic-cancer-risk-in-large-long-lived-afrotherians}{%
\subsection{Step-wise reduction of intrinsic cancer risk in large,
long-lived
Afrotherians}\label{step-wise-reduction-of-intrinsic-cancer-risk-in-large-long-lived-afrotherians}}

\begin{verbatim}
## Warning: TODO: Caption Fig 5
\end{verbatim}

\begin{verbatim}
## Warning: TODO: de-comment geom_tiplab in f5a! Commented currently for offline
## usage!
\end{verbatim}

\includegraphics{paper_PLOS_draft_files/figure-latex/Fig 5-1.pdf}

\begin{verbatim}
## 
## % Table created by stargazer v.5.2.2 by Marek Hlavac, Harvard University. E-mail: hlavac at fas.harvard.edu
## % Date and time: Wed, May 13, 2020 - 20:08:58
## \begin{table}[!htbp] \centering 
##   \caption{Phylogenetic Least Squares: ln(Lifespan) & ln(Body Size) Regression} 
##   \label{} 
## \begin{tabular}{@{\extracolsep{5pt}}lc} 
## \\[-1.8ex]\hline 
## \hline \\[-1.8ex] 
##  & \multicolumn{1}{c}{\textit{Dependent variable:}} \\ 
## \cline{2-2} 
## \\[-1.8ex] & Lifespan \\ 
## \hline \\[-1.8ex] 
##  lnSize & 0.100 \\ 
##   & (0.121) \\ 
##   & t = 0.826 \\ 
##   & p = 0.409 \\ 
##   Constant & 1.943 \\ 
##   & (1.385) \\ 
##   & t = 1.403 \\ 
##   & p = 0.161 \\ 
##  \hline \\[-1.8ex] 
## Observations & 28 \\ 
## Log Likelihood & $-$39.636 \\ 
## Akaike Inf. Crit. & 85.273 \\ 
## Bayesian Inf. Crit. & 89.047 \\ 
## \hline 
## \hline \\[-1.8ex] 
## \textit{Note:}  & \multicolumn{1}{r}{$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01} \\ 
## \end{tabular} 
## \end{table}
\end{verbatim}

>>>>

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Pervasive duplication of tumor suppressor genes preceded parallel
evolution of large bodied Atlantogenatans} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Juan Manuel Vazquez\textsuperscript{\getAff{University of Chicago}}\textsuperscript{*},
Vincent J Lynch\textsuperscript{\getAff{University at Buffalo, SUNY}}\\
\bigskip
\textbf{\getAff{University of Chicago}}Department of Human Genetics, 920 East 58th St, Chicago, IL, 60637\\
\textbf{\getAff{University at Buffalo, SUNY}}Department of Biological Sciences, 551 Cooke Hall, Buffalo NY, 14260\\
\bigskip
* Corresponding author: juanvazquez@uchicago.edu\\
\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
Cancer is an intrinsic disease of multicellular organisms. Within a
species, the size of an animal, - correlated with the individual's
number of cells - and its lifespan - correlated with increasing cellular
damage over time - positively correlate with the risk any individual has
to form tumors. Between species, however, we do not observe any
correlation between size, lifespan, and cancer, a phenomenon that
referred to as \emph{Peto's Paradox}. Elephants are a particularly
interesting member of this class of paradoxical animals, since they are
a set of large species deeply nested in a clade of smaller species,
indicating a recent gain of size. Recent work has identified several
individual cases of gene duplicates contributing to the increased cancer
resistance of elephants, which suggests that duplication of tumor
suppressor genes may play a more general role in mediating Peto's
Paradox by increasing cancer resistance in large, long-lived species. By
using a Reciprocal Best-Hit BLAT search approach, we investigated copy
numbers of all protein-coding genes in \emph{Atlantogenatan} genomes to
see if there is any correlation between the copy number of duplicates
and changes body size along the phylogenetic tree. From an initial set
of 18,011 protein-coding genes in hg38, we identified a median of 13,880
genes in \emph{Atlantogenatan} genomes, of which a median of 940 genes
are duplicated. We find that, just as body size fluctuates throughout
\emph{Atlantogenata}, genes involved in tumor suppressor pathways are
also duplicated throughout the phylogenetic tree. Extant species of
elephants, however, show active transcription of both canonical and
duplicated copies of tummor suppressors that duplicated prior to and
during their sudden increase in body size, suggesting that the
duplication of tumor suppressor genes facilitates the evolution of
increased body size by compensating for the increased cancer risk.

@@

\textbf{Reciprocal Best-Hit BLAT:} We developed a reciprocal best hit
BLAT (RBHB) pipeline to quickly identify homologs and estimate gene copy
numbers (\textbf{Figure 1A}). The Reciprocal Best Hit (RBH) search
strategy is conceptually straightforward: 1) Given a gene of interest
\(G_A\) in a query genome \(A\), one searches a target genome \(B\) for
all possible matches to \(G_A\); 2) For each of these hits, one then
performs the reciprocal search in the original query genome to identify
the highest-scoring hit; 3) A hit in genome \(B\) is defined as a
homolog of gene \(G_A\) if and only if the original gene \(G_A\) is the
top reciprocal search hit in genome \(A\). We selected BLAT {[}26{]} as
our algorithm of choice, as this algorithm is sensitive to highly
simliar (\textgreater{}90\% identity) sequences, thus identifying the
highest-confidence homologs while minimizing many-to-one mapping
problems when searching for multiple genes. RBH performs similar to
other more complex methods of orthology prediction, and is particularly
good at identifying incomplete genes that may be fragmented in low
quality/poor assembled regions of the genome {[}{\textbf{???}},27{]}.

\textbf{Effective Copy Number By Coverage:} In lower-quality genomes,
many genes are fragmented across multiple scaffolds, which results in
BLAT calling multiple hits when in reality there is only one gene. To
compensate for this, we came up with a novel statistic, Estimated Copy
Number by Coverage (ECNC), which averages the number of times we see
each nucleotides of a query sequence in a target genome over the total
number of nucleotides of the query sequence found overall in each target
genome (Supplementary Figure 1). This allows us to correct for genes
that have been fragmented across incomplete genomes, while also taking
into account missing sequences from the human query in the target
genome. Mathematically, this can be written as:\\
\[ ECNC = \frac{\sum_{n=1}^{l} C_n}{\sum_{n=1}^{l} bool(C_n)}\] where
\(n\) is a given nucleotide in the query, \(l\) is the total length of
the query, \(C_n\) is the number of instances that \(n\) is present
within a reciprocal best hit, and \(bool(C_n)\) is 1 if \(C_n > 0\) or 0
if \(C_n = 0\).

\textbf{RecSearch Pipeline} We created a custom Python pipeline for
automating RBHB searches between a single reference genome and multiple
target genomes using a list of query sequences from the reference
genome. For the query sequences in our search, we used the hg38 Proteome
provided by UniProt {[}28{]}, which is a comprehensive set of protein
sequences curated from a combination of predicted and validated protein
sequences generated by the UniProt Consortium. In order to refine our
search, we omitted protein sequences originating from long, noncoding
RNA loci (e.g.~LINC genes); poorly-studied genes from predicted open
reading frames (C-ORFs); and sequences with highly repetitive sequences
such as zinc fingers, protocadherins, and transposon-containing genes,
as these were prone to high levels of false positive hits. After
filtering out problematic protein queries, we then used our pipeline
(Figure 1A) to search for all copies of our 18011 query genes in
publicly available Afrotherian genomes, including African savannah
elephant (\emph{Loxodonta africana}: loxAfr3, loxAfr4, loxAfrC), African
forest elephant (\emph{Loxodonta cyclotis}: loxCycF), Asian Elephant
(\emph{Elephas maximus}: eleMaxD), Woolly Mammoth (\emph{Mammuthus
primigenius}: mamPriV), Colombian mammoth (\emph{Mammuthus columbi}:
mamColU), American mastodon (\emph{Mammut americanum}: mamAmeI), Rock
Hyrax (\emph{Procavia capensis}: proCap1, proCap2, proCap2\_HiC), West
Indian Manatee (\emph{Trichechus manatus latirostris}: triManLat1,
triManLat1\_HiC), Aardvark (\emph{Orycteropus afer}: oryAfe1,
oryAfe1\_HiC), Lesser Hedgehog Tenrec (\emph{Echinops telfairi}:
echTel2), Nine-banded armadillo (\emph{Dasypus novemcinctus}: dasNov3),
Hoffman's two-toed sloth (\emph{Choloepus hoffmannii}: choHof1, choHof2,
choHof2\_HiC), Cape golden mole (\emph{Chrysochloris asiatica}:
chrAsi1), and Cape elephant shrew (\emph{Elephantulus edwardii}:
eleEdw1). For many of these species, we covered multiple assemblies in
order to test the effects of assembly size and quality on our hits.

\textbf{Query protein criteria} To assemble our query list, we first
removed all unnamed genes from UP000005640. Next, we excluded genes from
downstream analyses for which assignment of homology was uncertain,
including uncharacterized ORFs (991), LOC (63), HLA genes (402),
replication dependent histones (72), odorant receptors (499), ribosomal
proteins (410), zinc finger transcription factors (1983), viral and
repetitive-element-associated proteins (82) and any protein described as
either ``Uncharacterized,'' ``Putative,'' or ``Fragment'' by UniProt in
UP000005640 (30724), leaving us with a final set of 37582 query protein
sequences, corresponding to 18011 genes.

\textbf{Duplication gene inclusion criteria} In order to condense
transcript-level hits into single gene loci, and to resolve many-to-one
genome mappings, we removed exons where transcripts from different genes
overlapped, and merged overlapping transcripts of the same gene into a
single gene locus call. The resulting gene-level copy number table was
then combined with the maximum ECNC values observed for each gene in
order to call gene duplications. We called a gene duplicated if its copy
number was two or more, and if the maximum ECNC value of all the gene
transcripts searched was 1.5 or greater; previous studies have shown
that incomplete duplications can encode functional genes, therefore
partial gene duplications were included provided they passed additional
inclusion criteria. The ECNC cut off of 1.5 was selected empirically, as
this value minimized the number of false positives seen in a test set of
genes and genomes. The results of our initial search are summarized in
Figure 1B. Overall, we identified 13880 genes across all species, or
77.1\% of our starting query genes.

\textbf{Genome Quality Assessment using CEGMA} In order to determine the
effect of genome quality on our results, we used the gVolante webserver
and CEGMA to assess the quality and completeness of the genome
{[}{\textbf{???}},{\textbf{??}}{]}. CEGMA was run using the default
settings for mammals (``Cut-off length for sequence statistics and
composition'' = 1;``CEGMA max intron length'' = 100000; ``CEGMA gene
flanks'' = 10000, ``Selected reference gene set'' = CVG). For each
genome, we generated a correlation matrix using the aforementioned
genome quality scores, and either the mean Copy Number or mean ECNC for
all hits in the genome.

@@

In order to determine the cancer risk \(K\) at each node, we first
needed to calcultate ancestral lifespans at each node. To do so, we used
a Phylogenetic Generalized Least-Square Regression (PGLS)
{[}{\textbf{???}},{\textbf{??}}{]} to calculate estimated ancestral
lifespans across \emph{Atlantogenata} using our estimates for body size
at each node.

Next, we used a simplified multistage cancer risk model for body size
\(D\) and lifespan \(t\): \(K \approx Dt^6\)
{[}{\textbf{???}},{\textbf{??}},{\textbf{??}}{]}. To determine the
change in \(K\) between nodes, we obtained the ratio between the cancer
risk \(K_{1}\) at any given node, and the cancer risk \(K_{2}\) at its
ancestral node, using the equation
\(\frac{K_{2}}{K_{1}} \approx \frac{D_{1}t_{1}^6}{D_{2}t_{2}^6}\).
Finally, to simplify comparisons, we calculated the fold change cancer
risk between a node and its ancestor as
\(\log_{2}(\frac{K_{2}}{K_{1}})\).

@@

\textbackslash{}begin\{figure\}

\includegraphics{paper_PLOS_draft_files/figure-latex/Figure 1-1}
\hfill{}

\textbackslash{}caption\{Figure 1: Body sizes rapidly and frequently
expand in Eutherians, especially in Atlantogenata. \textbf{A)} Tree of
Eutherian species, colored by ln(Body Size) and with branch lengths set
to the rate of change in body sizes, normalized by the square root of
the root branch. Atlantogenata is highlighted at the bottom. \textbf{B)}
Zoom-in of (A) on Atlantogenata. Silhuetes for the African Elephant,
West Indian Manatee, Cape Elephant Shrew, Lesser Hedgehog Tenrec, Cape
Golden Mole, Nine-Banded Armadillo, and Hoffman's Two-Toed Sloth are
colored by their extant body sizes, while clade labels are colored based
on the common ancestor's estimated body size\}\label{fig:Figure 1}
\textbackslash{}end\{figure\}

@@

Node & Size (log(g)) & 95\% CI (Low) & 95\% CI (High) & Rate
(sqrt)\tabularnewline

@@

To trace the evolutionary history of body mass and lifespan in
Afrotherians, we built a time-calibrated supertree of Eutherian mammals
combining 1,679 species from Bininda-Emonds et al {[}24{]} with a total
evidence Afrotherian phylogeny including 77 extant and fossil data from
39 extinct species {[}19{]}. Fossil data from extinct species were
included to ensure that ancestral state reconstructions of body mass in
Afrotherians were not biased by only including extant species, which can
lead to inaccurate reconstructions, for example, if lineages multiple
lineages evolved large body masses from a small bodied ancestor. We
jointly estimated rates of body mass evolution and reconstructed
ancestral states using a generalization of a Brownian model of character
evolution, which allows for occasional large jumps in traits (stable
model) and out performs standard Brownian motion and Ornstein-Uhlenbeck
models of character evolution {[}25{]}.

Similar to previous studies of Afrotherian body size {[}19,25{]}, we
found that the body mass of the Afrotherian ancestor was inferred to be
small (0.26kg, 95\% CI: 0.31-3.01kg) and that substantial accelerations
in the rate of body mass evolution occurred coincident with a 2.77x
increase \textbf{(Vinnie wrote 65× increase")} in body mass in the
stem-lineage of \emph{Pseudoungulata} (9.76kg) \textbf{(``Vinnie wrote
(17kg)'')}, a 1.57x increase \textbf{(``Vinnie wrote 1.5x increase'')}
in body mass in the stem-lineage of \emph{Paenungulata} (10.13kg)
\textbf{(``Vinnie wrote (25kg)'')}, a 1.52x increase \textbf{(``Vinnie
wrote 12× increase'')} in body mass in the stem-lineage of
\emph{Tehthytheria} (12.6kg) \textbf{(``Vinnie wrote (296kg)'')}, and a
1.22x increase \textbf{(``Vinnie wrote 14× increase'')} in body mass in
the stem-lineage of \emph{Proboscidea} (15.23kg) \textbf{(``Vinnie wrote
(4,100kg)'')} (Figure 1). The ancestral \emph{Hyracoidea} was inferred
to be relatively small 7.96kg-11.68kg \textbf{(Vinnie wrote
(2.86-15.71kg))}, and rate accelerations were coincident with
independent body mass increases in large hyraxes such as
\emph{Titanohyrax andrewsi} 12.97kg; 1.33x increase (67× increase in
body mass). While the body mass of the ancestral \emph{Sirenian} was
inferred to be large (61-656kg), a rate acceleration occurred coincident
with a 10× body mass increase in Stellar's sea cow. Rate accelerations
also occurred coincident with 36× body mass reduction in the
stem-lineage of the dwarf elephants \emph{Elephas}
(\emph{Palaeoloxodon}) \emph{falconeri} and \emph{Palaeoloxodon
cypriotes}. These data suggest that gigantism in \emph{Afrotherians}
evolved step-wise, from small to medium bodies in the
\emph{Pseudoungulata} stem-lineage, medium to large bodies in the
\emph{Tehthytherian} stem-lineage and extinct hyraxes, and from large to
exceptionally large bodies independently in the \emph{Proboscidean}
stem-lineage and Stellar's sea cow (Figure 1).

\hypertarget{pervasive-duplication-of-tumor-suppressors-during-the-origins-of-large-bodied-afrotheirans}{%
\subsection*{Pervasive duplication of tumor suppressors during the
origins of large bodied
Afrotheirans}\label{pervasive-duplication-of-tumor-suppressors-during-the-origins-of-large-bodied-afrotheirans}}
\addcontentsline{toc}{subsection}{Pervasive duplication of tumor
suppressors during the origins of large bodied Afrotheirans}
